---
title: 'Geog6300: Lab 6'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Regression #reference class day 11/14/2024
#univariate regression
#multivariate regression
#

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(tidyverse)
library(knitr)
library(tmap)

library(car)
library(lmtest)
library(Hmisc)
library(sjPlot)
library(corrplot)
```

**Overview:**
This lab focuses on regression techniques. You'll be analyzing the association of various physical and climatological characteristics in Australia with observations of several animals recorded on the citizen science app iNaturalist.

###Data and research questions###

Let's import the dataset.

```{r}
lab6_data<-st_read("data/aus_climate_inat.gpkg")
```

The dataset for this lab is a 1 decimal degree hexagon grid that has aggregate statistics for a number of variables:

* ndvi: NDVI/vegetation index values from Landsat data (via Google Earth Engine). These values range from -1 to 1, with higher values indicating more vegetation.
* maxtemp_00/20_med: Median maximum temperature (C) in 2000 or 2020 (data from SILO/Queensland government)
* mintemp_00/20_med: Median minimum temperature (C) in 2020 or 2020 (data from SILO/Queensland government)
* rain_00/20_sum: Total rainfall (mm) in 2000 or 2020 (data from SILO/Queensland government)
* pop_00/20: Total population in 2000 or 2020 (data from NASA's Gridded Population of the World)
* water_00/20_pct: Percentage of land covered by water at some point during the year in 2000 or 2020
* elev_med: Median elevation (meters) (data from the Shuttle Radar Topography Mission/NASA)

There are also observation counts from iNaturalist for several distinctively Australian animal species: the central bearded dragon, the common emu, the red kangaroo, the agile wallaby, the laughing kookaburra, the wombat, the koala, and the platypus.

Our primary research question is how the climatological/physical variables in our dataset are predictive of the NDVI value. We will build models for 2020 as well as the change from 2000 to 2020. The second is referred to as a "first difference" model and can sometimes be more useful for identifying causal mechanisms.

###Part 1: Analysis of 2020 data###

We will start by looking at data for 2020. 

**Question 1** _Create histograms for NDVI, max temp., min temp., rain, and population, and water in 2020 as well as elevation. Based on these graphs, assess the normality of these variables._

```{r}
#Code goes here.
kable(head(lab6_data))
#hist(lab6_data) #error
colnames(lab6_data)

lab6_data_numeric <- lab6_data %>%
  select(ndvi_20_med, maxtemp_20_med, mintemp_20_med, rain_20_sum, pop_20, water_20_pct, elev_med) #make sure numeric #ndvi does not exist as a column.

#lab6_data_numeric <- lab6_data %>%
  #select(ndvi_20_med, maxtemp_00_20_med, mintemp_00_20_med, rain_00_20_sum, pop_00_20, water_00_20_pct, elev_med) #make sure numeric #ndvi does not exist as a column.

#NDVI 2020
hist(lab6_data_numeric$ndvi_20_med)  #skewed

#Max Temp 2020
hist(lab6_data_numeric$maxtemp_20_med) #skewed

#Min Temperature 2020
hist(lab6_data_numeric$mintemp_20_med) #skewed

#Rain 2020
hist(lab6_data_numeric$rain_20_sum) #skewed

# Population 2020
hist(lab6_data_numeric$pop_20) #skewed

# Water Percentage 2020
hist(lab6_data_numeric$water_20_pct) #skewed

# Elevation
hist(lab6_data_numeric$elev_med) #skewed

```

{All of these variables look skewed (not normally distributed) based on their histograms.}

**Question 2** _Use tmap to map these same variables using Jenks natural breaks as the classification method. For an extra challenge, use `tmap_arrange` to plot all maps in a single figure._

```{r}
map_ndvi <- tm_shape(lab6_data) + # error correction - whole sf object needed to be passed in to tm_shape
  tm_borders() +  
  tm_fill("ndvi_20_med" , style = "jenks") #needed to specify the variable for the color, in this case "ndvi_20_med"

map_maxT <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("maxtemp_20_med" , style = "jenks") 

map_minT <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("mintemp_20_med" , style = "jenks")

map_rain <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("rain_20_sum" , style = "jenks")

map_pop <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("pop_20" , style = "jenks")

map_wat <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("water_20_pct" , style = "jenks")

map_elev <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("elev_med" , style = "jenks")


tmap_arrange(map_ndvi, map_maxT, map_minT, map_rain, map_pop, map_wat, map_elev )
```


**Question 3** _Based on the maps from question 3, summarise major patterns you see in the spatial distribution of these data from any of your variables of interest. How do they appear to be associated with the NDVI variable?_

{Rainfall is concentrated along the north and western coasts which matches the pattern of the NDVI.}

**Question 4** _Create univariate models for each of the variables listed in question 1, with NDVI in 2020 as the dependent variable. Print a summary of each model. Write a summary of those results that indicates the direction, magnitude, and significance for each model coefficient._

```{r}

mod_maxT<-lm(ndvi_20_med~maxtemp_20_med,data=lab6_data)
mod_minT<-lm(ndvi_20_med~mintemp_20_med,data=lab6_data)
mod_rain<-lm(ndvi_20_med~rain_20_sum,data=lab6_data)

mod_pop<-lm(ndvi_20_med~pop_20,data=lab6_data)
mod_wat<-lm(ndvi_20_med~water_20_pct,data=lab6_data)
mod_elev<-lm(ndvi_20_med~elev_med,data=lab6_data)



summary(mod_maxT)
summary(mod_minT)
summary(mod_rain)

summary(mod_pop)
summary(mod_wat)
summary(mod_elev)


```

{As elevation increases by 1 unit, the NDVI increases by 1.787e-04 which is a significant relationship
As the watercover increases by 1 unit, the NDVI decreases by -0.178263 which is not a significant relationship
As the population increases by 1 unit, the NDVI increases by 1.500e-06 which is a significant relationship but less significant than the other factors.
As rain fall increases by 1 unit, the NDVI increases by 9.124e-07 which is a significant relationship
As mintemp increases by 1 unit, the NDVI decreases by -0.012282 which is a significant relationship
As maxtemp increases by 1 unit, the NDVI decreases by -0.0130902 which is a significant relationship}

**Question 5** _Create a multivariate regression model with the variables of interest, choosing EITHER max or min temperature (but not both) You may also choose to leave out any variables that were insignificant in Q4. Use the univariate models as your guide. Call the results._

```{r}
model_all<-lm(ndvi_20_med~maxtemp_20_med + rain_20_sum + pop_20 + elev_med,data=lab6_data)
summary(model_all)


```

**Question 6** _Summarize the results of the multivariate model. What are the direction, magnitude, and significance of each coefficient? How did it change from the univariate models you created in Q4 (if at all)? What do the R2 and F-statistic values tell you about overall model fit?_

{As elevation increases by 1 unit, the NDVI increases by  1.233e-04 which is lower then the univariate result of 1.787e-04 and both are significant relationships
As the population increases by 1 unit, the NDVI increases by 2.862e-07 which is higher than the univariate result of 1.500e-06 and both are significant relationships but with two "stars" so less significant that the other factors.
As rain fall increases by 1 unit, the NDVI increases by 8.469e-07 which is lower than the univariate result of 9.124e-07 and both are significant relationships
As maxtemp increases by 1 unit, the NDVI decreases by -1.170e-02 which is higher in magnitude than -0.0130902 and both are significant relationships

The F-statistic is an analysis of variance and it indicates whether or not the r2 is significant. The adjusted R2 is 0.6376 (so 63% of the change in NDVI is explained by the variables in question) and the f-statistic is 315.5 which is a high value so it indicates that the r2 is reliable. Based on these results, this model is a good "fit" meaning that the independent variables do a good job of explaining the change in NDVI.  }

**Question 7** _Use a histogram and a map to assess the normality of residuals and any spatial autocorrelation. Summarise any notable patterns that you see._

```{r}

lab6_data$residuals<-residuals(model_all)

hist(lab6_data$residuals)

map_res <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("residuals" , style = "jenks")
map_res 

```

{the most notable residuals (most extreme) seemed scattered allong the coast}

**Question 8** _Assess any issues with multicollinearity or heteroskedastity in this model using the techniques shown in class. Run the appropriate tests and explain what their results show you._


```{r}

vif(model_all) #the value should be under 5 for low multicollinearity.

bptest(model_all) #skibbity heteroskedasticity # if the p-value is significant, I would reject the null that the data is NOT.


```

{The vif value was under 5 so I do not suspect high multicollinearity. The P value is significant at p-value < 2.2e-16, so I reject the null hypothesis and conclude that the data is heteroskedastic}

**Question 9** _How would you summarise the results of this model in a sentence or two? In addition, looking at the full model and your diagnostics, do you feel this is a model that provides meaningful results? Explain your answer._

```{r}
#In summary, the results of this model indicate that rainfall, Population and elevation have a positive impact on the increase of NDVI and max temperature has a negative relationship with NDVI. The model that I created which omits mintemp and watercover, is only somewhat changed from the full model, also, low multicolinearity was indicated so I would tend to trust this model to provide meaningful results. 
```


**Disclosure of assistance:** _Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?_

{A fellow student helped me debug question 4. Chatgpt helped me debug adding new columns in the challenge question. I couldn't remember where in the class material we talked about R2 is 0.6376 and the f-statistic so I looked that up onlilne and in our textbook}

**Lab reflection:** _How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What were the biggest things you learned by completing it?_

{I felt good about my work on this lab and gaining more comptency in R. It was a moderately hard lab and it was pretty. I REALLY appreciated using some climate data! The univariate versus multivariate analysis was particularly useful to help interpret, compare and cement an understanidng of these types of regression modeling. }


**Challenge question**

#Option 1
Create a first difference model. To do that, subtract the values in 2000 from the values in 2020 for each variable for which that is appropriate. Then create a new model similar to the one you created in question 5, but using these new variables showing the *change in values* over time. Call the results of the model, and interpret the results in the same ways you did above. Also chart and map the residuals to assess model error. Finally, write a short section that summarises what, if anything, this model tells you. 

```{r}
#model_all_New<-lm(ndvi_20_med-ndvi_00_med~maxtemp_20_med + rain_20_sum,data=lab6_data) do the subtraction seperately and rename!
#summary(model_all)

dif1_NDVI<-(lab6_data$ndvi_20_med-lab6_data$ndvi_00_med)
dif1_maxt<-(lab6_data$maxtemp_20_med-lab6_data$maxtemp_00_med)
dif1_mint<-(lab6_data$mintemp_20_med-lab6_data$mintemp_00_med)
dif1_rain<-(lab6_data$rain_20_sum-lab6_data$rain_00_sum)
dif1_pop<-(lab6_data$pop_20-lab6_data$pop_00)
dif1_wat<-(lab6_data$water_20_pct-lab6_data$water_00_pct)

#dif1_NDVI
#dif1_maxt
#dif1_mint
#dif1_rain
#dif1_pop
#dif1_wat

Model_all_dif1<-lm(dif1_NDVI~dif1_maxt + dif1_mint + dif1_rain + dif1_pop + dif1_wat, data=lab6_data)
summary(Model_all_dif1)

#according to the first difference model, if correctly done, the only significant coefficients were the difference in rain and difference in min temp. I think this means that the other coefficients did not have significant change between 2000 and 2020. According to The min temp in 2000 was 

#NDVI first difference
hist(dif1_NDVI)  #fairly normal looking

#Max Temp first Dif
hist(dif1_maxt) #normalish

#Min Temperature first dif
hist(dif1_mint) #skewed

#Rain first dif
hist(dif1_rain) #normalish

# Population first dif
hist(dif1_pop) #skewed

# Water Percentage first dif
hist(dif1_wat) #normalish



```

```{r}

lab6_data$dif1_NDVI <- lab6_data$ndvi_20_med - lab6_data$ndvi_00_med
lab6_data$dif1_maxt <- lab6_data$maxtemp_20_med - lab6_data$maxtemp_00_med
lab6_data$dif1_mint <- lab6_data$mintemp_20_med - lab6_data$mintemp_00_med
lab6_data$dif1_rain <- lab6_data$rain_20_sum - lab6_data$rain_00_sum
lab6_data$dif1_pop <- lab6_data$pop_20 - lab6_data$pop_00
lab6_data$dif1_wat <- lab6_data$water_20_pct - lab6_data$water_00_pct

map_ndvi <- tm_shape(lab6_data) + # error correction - whole sf object needed to be passed in to tm_shape
  tm_borders() +  
  tm_fill("dif1_NDVI" , style = "jenks") #needed to specify the variable for the color, in this case "ndvi_20_med"

map_maxT <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("dif1_maxt" , style = "jenks") 

map_minT <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("dif1_mint" , style = "jenks")

map_rain <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("dif1_rain" , style = "jenks")

map_pop <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("dif1_pop" , style = "jenks")

map_wat <- tm_shape(lab6_data) + 
  tm_borders() +  
  tm_fill("dif1_wat" , style = "jenks")



tmap_arrange(map_ndvi, map_maxT, map_minT , map_rain, map_pop, map_wat )

```

{summary: The results showed predominantly negative change (indicating decrease over the time period 2000-2020) for NDVI, and rain. There were patterns of primarily increase in Max Temperature, and increase in Min temperature values concentrated accross western Australia. Population showed no decreases over Australia, and there were notable increases in population along the southeast coast. Percent water cover showed increase in some areas and decrease in others but with no distinct pattern.}

#Option 2
The animal data included in this dataset is an example of count data, and usually we would use a Poisson or similar model for that purpose. Let's try it with regular OLS regression though. Create two regression models to assess how the counts of two different animals (say, koalas and emus) are associated with at least three of the environmental/climatological variables given above. Be sure to use the same independent variables in each model. Interpret the results of each model and then explain the importance of any differences in the model coefficients between them, focusing on direction, magnitude, and significance.

```{r}



```

